---
title: "Multiple Coastal Stressors - Environmental Data"
author: "Gabriel Barrett"
slug: "EnvironmentalData"
date: "2022-22-08"
author_email: "gabriel_barrett@uri.edu"
author_github: "Gabriel-A-Barrett"
description: "Environmental data retrieval and curation for mutliple coastal stressors NSF project"
output: 
  html_document:
    self_contained: no
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
library(dplyr)
library(ggplot2)
library(dataRetrieval)
library(readxl)
library(leaflet)
library(viridis)
library(readxl)
library(lubridate)
source("D:/Puritz_Lab/NSF/NSF_EnvironmentalData/scripts/bin/Rfunctions.R")

#knit_hooks$set(plot=function(x, options) {
#  paste0('<figure><img src="/',options$fig.path,options$label,
#         "-", options$fig.cur,".",options$fig.ext,'" title = "',
#         options$fig.cap, '" alt = "', options$alt.text,'" ></figure>')
#})


opts_chunk$set(
    echo=TRUE,
    eval=TRUE,
    root.dir = "D:/Puritz_Lab/NSF/NSF_EnvironmentalData/",
    base.dir = "D:/Puritz_Lab/NSF/NSF_EnvironmentalData/",
    fig.path="assets/",
    fig.cap="TODO",
    alt.text="TODO",
    fig.width=7,
    fig.height=5,
    class="",
    message=FALSE,
    warnings=FALSE
)
```


# All Site Information

leaflet interactive map

```{r leaflet, echo=FALSE,eval=TRUE,warning=FALSE, fig.cap="Fig. 1 leaflet map showing site positions with popups containing site metadata"}

env <- read.delim("D:/Puritz_Lab/NSF/NSF_EnvironmentalData/data/NSF_EnvironmentalData.tsv",sep="\t") %>% 
  group_by(stn_id) %>%
  dplyr::summarise(
    stn_id=stn_id,
    location=location,
    avg_temp=as.numeric(round(mean(TEMP,na.rm=TRUE),digits=2)),
    avg_do= as.numeric(round(mean(DO,na.rm=TRUE),digits=2)),
    avg_ph= as.numeric(round(mean(PH,na.rm=TRUE),digits=2)),
    avg_sal= as.numeric(round(mean(SAL,na.rm=TRUE),digits=2)),
    depth=NA,
    lat=as.numeric(lat),
    long=as.numeric(long),
    coalition,
    begin_date=min(as.POSIXct(date, format='%Y-%m-%d')),
    end_date=max(as.POSIXct(date, format='%Y-%m-%d')),
    n=n()
    ) %>%
  distinct(stn_id,.keep_all = TRUE) %>% filter(n > 1)


waste <- read.delim("D:/Puritz_Lab/NSF/NSF_EnvironmentalData/data/wastewater_discarge_permits.tsv",sep="\t")

markers <- ~paste(
               "<br>Site Code: ", stn_id,
               "<br>Location: ", location,
               "<br>Mean Temperature: ",avg_temp,
               "<br>Mean Salinity: ", avg_sal,
               "<br>Mean DO: ", avg_do,
               "<br>Mean pH: ", avg_ph,
               "<br>Deployment: ", begin_date," to ", end_date,
               "<br>observations: ", n)

leaflet(width="100%") %>% addTiles() %>% 
  # USGS
  addMarkers(data = env %>% filter(coalition=="WQP"), lng = ~long, lat = ~lat,
             popup=~paste(
              "<br>Site Code: ", stn_id,
              "<br>Location: ", location,
              "<br>Mean Temperature: ",avg_temp,
              "<br>Mean Salinity: ", avg_sal,
              "<br>Mean DO: ", avg_do,
              "<br>Mean pH: ", avg_ph,
              "<br>Deployment: ", begin_date,"to", end_date,
              "<br>observations: ", n),
             clusterOptions = markerClusterOptions(), group = "WQP") %>%
  # Buzzards
  addMarkers(data = env %>% filter(coalition=="buzzards"), lng = ~long, lat = ~lat,
             popup=~paste(
               "<br>Site Code: ", stn_id,
               "<br>Location: ", location,
               "<br>Mean Temperature: ",avg_temp,
               "<br>Mean Salinity: ", avg_sal,
               "<br>Mean DO: ", avg_do,
               "<br>Mean pH: ", avg_ph,
               "<br>Deployment: ", begin_date," to ", end_date,
               "<br>observations: ", n),
             clusterOptions = markerClusterOptions(), group = "Buzzards Bay Coalition") %>%
  # Cheseapeak
  addMarkers(data = env %>% filter(coalition=="eyesonthebay"), lng = ~long, lat = ~lat,
             popup=~paste(
               "<br>Site Code: ", stn_id,
               "<br>Location: ", location,
               "<br>Mean Temperature: ",avg_temp,
               "<br>Mean Salinity: ", avg_sal,
               "<br>Mean DO: ", avg_do,
               "<br>Mean pH: ", avg_ph,
               "<br>Deployment: ", begin_date," to ", end_date,
               "<br>observations: ", n),
             clusterOptions = markerClusterOptions(), group = "Eyes On the Bay") %>%
  # Rhode Island DOH
  addMarkers(data = env %>% filter(coalition=="ridoh"), lng = ~long, lat = ~lat,
             popup=~paste(
               "<br>Site Code: ", stn_id,
               "<br>Location: ", location,
               "<br>Mean Temperature: ",avg_temp,
               "<br>Mean Salinity: ", avg_sal,
               "<br>Mean DO: ", avg_do,
               "<br>Mean pH: ", avg_ph,
               "<br>Deployment: ", begin_date," to ", end_date,
               "<br>observations: ", n),
             clusterOptions = markerClusterOptions(), group = "RI DOH") %>%
  # WasteWater
  addMarkers(data=waste,lng=~LONGITUDE,lat=~LATITUDE,
             popup=~paste(
               "<br>PGM System ID: ", PGM_SYS_ID,
               "<br>Primary Name: ", PRIMARY_NAME,
               "<br>Address: ", LOCATION_ADDRESS,
               "<br>More Information: <a href = ",FACILITY_URL,"> Link </a>"),
             clusterOptions = markerClusterOptions(), group = "Water Discharge Facilities") %>% 
  addLayersControl(overlayGroups = c("WQP","Buzzards Bay Coalition","Eyes On the Bay","RI DOH","Water Discharge Facilities"),
                   options= layersControlOptions(collapsed = FALSE)) %>% hideGroup("Water Discharge Facilities")

```

# Buzzards Bay Water Quality Coalition

Water Quality data based Buzzards Bay Water Quality Coalition is placed with WHOI archives: https://darchive.mblwhoilibrary.org/handle/1912/25762

```{r buzzards curation, echo=TRUE, eval=TRUE,warning=FALSE}

qc_exclude <- c(2,4,6,7,8,9)

buz <- read_excel("D:/Puritz_Lab/NSF/NSF_EnvironmentalData/data/bbcdata1992to2020-ver07May2021.xlsx",sheet=2) %>% 
  
  left_join(.,read_excel("D:/Puritz_Lab/NSF/NSF_EnvironmentalData/data/bbcdata1992to2020-ver07May2021.xlsx",sheet = 6,
                         col_names = T,skip=1),by="STN_ID") %>% 
  
  filter(
    !(TEMP_QC %in% qc_exclude),
    !(SAL_QC %in% qc_exclude),
    !(DO_QC %in% qc_exclude),
    !(PH_QC %in% qc_exclude),
    !(TIME_QC %in% qc_exclude),
    !(is.na(STN_ID))) %>%
    
  mutate(
    time = format(as.POSIXct(TIME, format='%Y-%m-%d %H:%M:%S'),format='%H:%M:%S'),
    time_stamp = as.POSIXct(paste(SAMP_DATE,time,sep=" "),format = '%Y-%m-%d %H:%M:%S'),
    coalition="buzzards")

# This step transforms datasets so that they are able to be rbinded 
buz_curated <- buz %>%
  select(stn_id=STN_ID,
         location=WQI_Area,
         town=Town,
         TEMP=TEMP_C,
         DO=DO_MGL,
         PH=PH,
         SAL=SAL_FIELD,
         depth=TOTDEP_M,
         lat=LATITUDE,
         long=LONGITUDE,
         coalition,time_stamp)
```

```{r dataframe, echo=FALSE, eval=TRUE,warning=FALSE, fig.cap="**Table 1.** Buzzards Bay Water Quality Coalition example dataframe"}
head(buz) %>% kbl(caption = paste("Buzzards Bay Coalition (n = ",nrow(buz),")")) %>% kable_classic(full_width=T)
```

**Table 1.** Buzzards Bay Water Quality Coalition example dataframe

Went on to exclude data based on quality control codes provided...

```{r buzzardsbay, echo = FALSE, eval = TRUE, fig.cap = "**Table 2.** quality control codes excluded from dataset"}
qc_df <- data.frame(
    "code" = c(1,2,3,4,6,7,8,9,10,11,12),
    "description" = c("centrigade recalculate from Fahrenheit","value needs to be reviewed for accuracy","value reviewed then confirmed, rerun, or revised","flagged as unusual value, with comments about possible reasons if known","substitute value with comment for details","data below method detection limit","value highly anomalous, although laboratory measurments were validated, sample may have been compromised during collection","value rejected","laboratory salinity substituded for missing field salinity","field salinity restored to original value for 1992-1995 for samples that had a correction factor applied at the time","time of D.O. collection too late for D.O. value to be used in buzzards bay coalition health index calculations"),
    "include_exclude" = c("include","exclude","include","exclude","exclude","exclude","exclude","exclude","include","include","include"))

qc_df %>% kbl(caption = "A Tale of Exclusion") %>% kable_classic(full_width=F)
```

**Table 2.** quality control codes excluded from dataset

### Time Series

```{r buzzardsbay_timeseries, echo = FALSE, eval = TRUE, fig.width=20, fig.height=10, fig.align = 'center', warning=FALSE, message=FALSE, fig.cap="**Fig. 1** Dissolved oxygen in mg/L across 4 sites w/in Buzzards Bay, demonstrating seasonal sampling", alt.text="capturing 4 sites with highest observation counts and DO taken as average across overlapping stations"}

sites <- c("Wareham River Inner","New Bedford Harbor Inner","Westport River East Inner","Mattapoisett Harbor Inner")
sub_buz <- buz %>% 
  select(WQI_Area, DO_MGL, time_stamp) %>% 
  filter(WQI_Area %in% sites)

EnvTimeSeriesPoints(data = sub_buz, x = time_stamp, y = DO_MGL, color = WQI_Area)
```

# Chesapeak Bay: "eyes on the bay"

Water quality data from "Eyes on the Bay" data query portal. Data quality control was done prior to uploading on the eyes on the bay website. Data was converted to wide format and subsetted to surface water samples. 

```{r chesepeak bay curation, echo = TRUE, eval = TRUE, messages=FALSE,warning=FALSE}
ches <- read.table("D:/Puritz_Lab/NSF/NSF_EnvironmentalData/data/waterquality_do_temp_sal_ph_chesapeakbay_Jan1990_Aug2022.txt",
                   sep="\t",header=T,
                   nrow=length(count.fields("D:/Puritz_Lab/NSF/NSF_EnvironmentalData/data/waterquality_do_temp_sal_ph_chesapeakbay_Jan1990_Aug2022.txt")) - 1, 
                   na.strings = "NA", fill = TRUE) %>% 
  
  left_join(.,read.delim("D:/Puritz_Lab/NSF/NSF_EnvironmentalData/data/WaterQualityStationHUC8_ChesapeakebayDataHub.txt",
                         sep="\t",header=TRUE), by="Station") %>% 
  filter(Layer == "S ") %>%
  
  mutate(row=row_number(),
         time_stamp=as.POSIXct(paste(SampleDate,SampleTime,sep=" "),format="%m/%d/%Y %H:%M:%S"),
         coalition="eyesonthebay") %>%
  
  pivot_wider(.,names_from = Parameter, values_from = MeasureValue) 


ches_curated <- ches %>% 
  select(stn_id=Station,
         location=StationDescription,
         town=CountyCity,
         TEMP=WTEMP,
         DO,
         PH,
         SAL=SALINITY,
         depth=TotalDepth,
         lat=Latitude.x,long=Longitude.x,
         coalition,time_stamp)
```

```{r chesepeak bay df, echo = FALSE, eval = TRUE, messages=FALSE,warning=FALSE}
head(ches) %>% kbl(caption = paste("Keep Your Eyes On the Bay (n=",nrow(ches_curated),")")) %>% kable_classic(full_width=F)
```

### Time Series

```{r chesepeak bay time series, echo=FALSE, eval = TRUE, message=FALSE, warning=FALSE, fig.width=25, fig.height=10, fig.cap="**Fig.2** Dissolved oxygen in mg/L across across 4 sites",alt.text="capturing 4 sites with highest observation counts and DO taken as average across overlapping sites",fig.align='center'}

sites <- c("CB5.1","LE1.4","CB3.3C","CB5.1W")

sub_ches <- ches %>% filter(Station %in% sites)

EnvTimeSeriesPoints(data = sub_ches, x = time_stamp, y = DO, color = StationDescription)
```


# Water Quality Portal (WQP)

Data was queryed via https://www.waterqualitydata.us/ 

Search was narrowed based on state, county, site type == "estuary", sample media == "Water (NWIS,STEWARDS,STORET)", characteristic group == "Inorganic, Minor, Metals", and data profiles == "Sample Results (narrow)"

```{r Water Quality Portal curation, echo = TRUE, eval = TRUE, messages=FALSE,warning=FALSE}

variables <- c("Dissolved oxygen (DO)", "Oxygen", "Temperature, water", 'Salinity', "pH") # Measurments We Want

fairfield <- read.csv("D:/Puritz_Lab/NSF/NSF_EnvironmentalData/data/WQP_CT_middlesex.csv",na.strings=c("","NA")) %>% 
  filter(CharacteristicName %in% variables) %>% 
  mutate(time_stamp=as.POSIXct(ActivityStartDate,format="%Y-%M-%d"))

sites <- fairfield %>% 
  distinct(MonitoringLocationIdentifier,.keep_all = TRUE) %>%
  group_by(MonitoringLocationIdentifier) %>%
  dplyr::mutate(end_date=max(time_stamp)) %>% 
  #dplyr::filter(end_date >= "2015-01-01") %>% 
  dplyr::select(site=MonitoringLocationIdentifier,end_date)

siteLoadedList <- lapply(sites$site,function(i){whatWQPdata(siteid=i)}) # One line information pertaining to sites activity, location, ect.
sitemeta <- do.call(rbind,siteLoadedList) # combine into one df

# Place Curated datasets within subsets list

curated_fairfield <- fairfield %>% 
  dplyr::group_by(MonitoringLocationIdentifier) %>%
  dplyr::mutate(row=dplyr::row_number(),
          result = as.numeric(ResultMeasureValue),
          datatype = gsub("([A-Za-z]+).*", "\\1",CharacteristicName),
         
         time_stamp=as.POSIXct(ActivityStartDate,format="%Y-%M-%d")) %>%
  
  # Remove unwanted units 
  filter(!(ResultMeasure.MeasureUnitCode == "ppth" || ResultMeasure.MeasureUnitCode == "deg F" || is.na(ResultMeasure.MeasureUnitCode) == TRUE || ResultMeasure.MeasureUnitCode == "%" || ResultMeasure.MeasureUnitCode == "ppm")) %>%
  
  # Must subset or else pivot wider does not fully combine Results into one row but multiple in different columns
  select(MonitoringLocationIdentifier,datatype,time_stamp,result) %>%
  
  pivot_wider(.,names_from="datatype",values_from="result", values_fn = ~mean(.x, na.rm = TRUE)) %>% 
  
  # Get meta information
  left_join(.,sitemeta,by="MonitoringLocationIdentifier") %>%
  
  # Final DataFrame Structure
  summarise(
    stn_id=MonitoringLocationIdentifier,
    location=MonitoringLocationName,
    town=CountyName,
    TEMP=as.numeric(Temperature),
    DO=as.numeric(Dissolved),
    PH=as.numeric(pH),
    SAL=as.numeric(Salinity),
    depth=NA,
    lat=lat,
    long=lon,
    coalition="WQP",
    time_stamp=time_stamp)

```

```{r fairfield WQP Query df, echo = FALSE, eval = TRUE, messages=FALSE,warning=FALSE}
head(fairfield) %>% kbl(caption = "Water Quality Portal Fairfield County, CT") %>% kable_classic(full_width=F)
```

```{r fairfield WQP time series, echo=FALSE, eval = TRUE, message=FALSE, warning=FALSE, fig.width=25, fig.height=10, fig.cap="**Fig.2** Dissolved oxygen in mg/L across across 4 sites",alt.text="capturing 4 sites with highest observation counts and DO taken as average across overlapping sites",fig.align='center'}

top <- curated_fairfield %>% group_by(stn_id) %>% summarise(n=n()) %>% arrange(desc(n)) %>% slice(1:4)
fairfieldsites <- top$stn_id 
sub_fairfield <- curated_fairfield %>% filter(stn_id %in% fairfieldsites)
EnvTimeSeriesPoints(data = sub_fairfield, x = time_stamp, y = DO, color = location)
```

# RIDOH
obtained dataset via emailing eherron@uri.edu. Values were obtained by taking the average across sampling depths and sampling times per day. 

```{r rhode island DOH, echo = TRUE, eval = TRUE, messages=FALSE,warning=FALSE}
ri_parametercodes <- read_xlsx("D:/Puritz_Lab/NSF/NSF_EnvironmentalData/data/RIDOH_Supporting_Files.xlsx",sheet=2) %>% 
  mutate(ParameterName = gsub("\\ -.*","",ParameterName)) 

ri_sites <- read_xlsx("D:/Puritz_Lab/NSF/NSF_EnvironmentalData/data/RIDOH_Supporting_Files.xlsx",sheet=5)

ridoh <- read.delim("D:/Puritz_Lab/NSF/NSF_EnvironmentalData/data/RIDOH_1988_2020.csv",sep = ',') %>% # description - code; removing description
  
  mutate(parameter_code = gsub(".* - ","",Parameter..),
         date = mdy(Date.of.Sample),
         time = hms(Time), # Some strings are missing time, only evaluated based date and avg across time for a given day
         # time_stamp = mdy_hms(paste0(Date.of.Sample," ",Time)),
         measured_value = as.numeric(Concentration),
         id = row_number()) %>% 
  
  filter(parameter_code %in% c("00300","00400","00480","00011"), # dissolved oxygen = 00300, pH = 00400, salinity = 00480, temperature = 00011
         !(Qualifier.Code %in% c("N","U","I","V")))

ridoh_curated <- ridoh %>%
  
  group_by(date, WW.ID, parameter_code) %>% # pooled different depths and samples from the same day
  
  summarise(avg_measuredvalue = mean(Concentration,na.rm=TRUE)) %>% # Took the average across these avg. depths and samples taken that day
  
  pivot_wider(names_from=parameter_code, values_from = avg_measuredvalue, names_prefix = "env_") %>% 
  
  left_join(.,ri_sites, by=c("WW.ID"="WW_Station")) %>%
  
  ungroup() %>%
  
  summarise(
    stn_id=WW.ID,
    location=Site_DESCR,
    town=Town,
    TEMP=env_00011,
    DO=env_00300,
    PH=env_00400,
    SAL=env_00480,
    depth=NA,
    lat=LAT_DD,
    long=LON_DD,
    coalition="ridoh",
    time_stamp=date)
```


```{r ridoh df, echo = FALSE, eval = TRUE, messages=FALSE,warning=FALSE}
head(ridoh) %>% kbl(caption = "Rhode Island Department of Health (DOH)") %>% kable_classic(full_width=F)
```

Values Were exclude based on being tagged for the below qualifier codes. If a record is absent from a qualifier code that it was assumed to be reflect accurate reading. 

```{r rhode island doh qualifier codes, echo = TRUE, eval = TRUE, messages=FALSE,warning=FALSE}
qualifier_codes <- data.frame(
  "Qualifier Code" = c("N","U","I","V","Z"),
  "Description" = c("No Results (see comments)","Indicates that the compound was analyzed but not detected. This code
shall be used to indicate that the lab value reported is less than the Method Detection Limit and is reported for informational purposes.","Value reported is equal to or greater than the Method Detection Limit but less
than the Quantitation Level (Reporting Level). Data shall be deemed invalid.","Analyte was detected in both the sample and the associated method blank.","Actual value is greater than the value reported. For bacteria analyses: Too many colonies were present (too numerous to count TNTC - membrane filtration) or all wells positive (IDEXX). Other analyses - off scale."), 
"include_exclude" = c("exclude","exclude","exclude","exclude","include"))

qualifier_codes %>% kbl(caption = "A Tale of Exclusion") %>% kable_classic(full_width=F)
```
# Water Treatment Facilities

obtained from epa [data portal](https://edg.epa.gov/metadata/catalog/search/resource/details.page?uuid=%7BD7F5EE9A-248F-43FA-85F4-CCFC1181BF25%7D). Points of National Pollutant Discharge Elimination System (NPDES) and Permit Compliance Systems (PCS) programs as reported in Integrated Compliance Information Systems (ICIS). This dataset is updated weekly. 

included only public facilities

```{r Water Treatment Facilities, echo = FALSE, eval = TRUE, messages=FALSE,warning=FALSE}
head(waste) %>% kbl(caption = "Wastewater Discharge Facilities & Sewers & Water Quality Control Facilties") %>% kable_classic(full_width=F)
```
